From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-23 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Incorporated changes from 2-22 meeting

In the continuous case, when $\mathcal{S}$ is a manifold, we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. We will call a set of quantities $\xi^a$ \textit{state variables} provided that they fully identify a state, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates'. However, we will have multiple manifolds and, to avoid confusion, we will call `state variables' the coordinates of the state space.} We define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ as the pullback of the distribution $\rho_c$ by the inverse of a coordinate map:  $\rho_{c,\xi} \equiv (\rho_c \circ \xi^{-1})$. %To represent the distribution over state variables, we define coordinate charts $(U, \xi)$ over $\mathcal{S}$. Likewise, the distribution $\rho_c$ is the pullback of $\rho^{\ast}_c$ by a coordinate map: $\rho_c (s) = (\xi^{\ast} \rho^{\ast}_c)(s) = \rho^{\ast}_c (\xi (s))$.

The distribution $\rho_c$ is a function of the particle states $s \in \mathcal{S}$, while the density $\rho_{c, \xi}$ is a function of the state variables $\xi^a \in \mathbb{R}^n$. A coordinate map $\xi$ lets us express the state $s$ as a function of the coordinates $\xi^a$. Consequently, we require that $\mu(U) = \int_U \rho_c(s) d\mathcal{S} = \int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a$.\footnote{Indeed, using the Lebesgue measure $\lambda$ on $\mathbb{R}^n$---and assuming that $\xi^{-1}$: $\mathbb{R}^n \to \mathcal{S}$ is measurable---we can define the measure $d\mathcal{S}$ by the pushforward measure $\xi^{-1}_{\ast}(\lambda)(U) \equiv \lambda(\xi(U))$.} Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$. Hence, $\rho_{c, \xi}$ will transform as a scalar under change of variables. Thus, $\int_U \rho_c(s) d\mathcal{S}$ is invariant under coordinate changes $\xi$ to $\hat{\xi}$. As a function of the state variables $\xi^a$, $\rho^{\ast}_c$ transforms according to the Jacobian for this change of variables: $\rho^{\ast}_c (\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c (\hat{\xi}^b)$. Since $\mu(U) = \int_{\xi(U)} \rho^{\ast}_c (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho^{\ast}_c (\hat{\xi}^b) d \hat{\xi}^b$ remains invariant under this coordinate change, the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless.\footnote{Formally, we can express this argument as a chain of equalities: $\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$, where $\rho^{\ast}_c(\xi^a)$ and $\rho^{\ast}_c(\hat{\xi}^b)$ are the same density expressed over different state variables and $s(\xi^a) \equiv \xi^{-1}(\xi^a)$ and $s(\hat{\xi}^b)\equiv \hat{\xi}^{-1}(\hat{\xi}^b)$ represent the same particle state as expressed by these different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through their labeling of a particle state $s$.} 
%$\rho^{\ast}_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho^{\ast}_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho^{\ast}_c(\hat\xi^b)$

We will now show that in order for the distribution $\rho_c$ to satisfy these two constraints, the classical phase space must be at least a symplectic manifold. In other words, having symplectic manifold structure is a necessary condition for the distribution over particle states to transform as a scalar function and for its pullback $\rho^{\ast}_c := \rho_c \circ \xi^{-1}$ to transform as a density.

%In other words, while we can change the units used by the state variables $\xi^a$ to identify the particle states, the units for $\rho^{\ast}_c$ must remain the same so that the density does not change its numeric value

@@

The units we choose for $\omega$ are really the units we choose to count states over an independent degree of freedom. This agrees with the practice in statistical mechanics. The density $\rho_c$ [XXX: distribution or density?], when normalized, is given in terms of the inverse of the units for $\omega$. In accordance with the standard conventions in quantum mechanics and statistical mechanics, the $k_i$ are given in terms of inverse units of the respective $q^i$, such that $dq^i \wedge dk_i$ is a pure number. We then use $\hbar$ to define the units for state counting and therefore set $\omega = \hbar dq^i \wedge dk_i = dq^i \wedge dp_i$ where $p_i = \hbar k_i$.
% % the `motivation' section of the wiki page on symplectic manifolds provides a different style of derivation for the 2-form being non-degenerate and closed.

@@

This completes our central argument for privileging Hamiltonian mechanics over Lagrangian mechanics. We have seen that Lagrangian systems are a proper subset of Hamiltonian systems, specifically those Hamiltonian systems that also satisfy the motion equivalence condition. In the remainder of this essay, we provide further reasons for privileging Hamiltonian mechanics. The next section describes how Hamiltonian mechanics provides the natural setting for the invariant density, describing the matter content of the theory [XXX: matter content of the theory seems like an odd phrase]. Section~\ref{fundamental} argues that the invariant phase space density is more fundamental than the state variables, since conjugate momentum is a gauge-dependent quantity. We then turn in Section~\ref{Lagrangian} to provide a geometric interpretation of the principle of stationary action within Hamiltonian mechanics. Our construction makes manifest the gauge redundancy of the Lagrangian, providing a further reason to privilege the Hamiltonian. Finally, we respond to Curiel's arguments against the fundamentality of Hamiltonian mechanics.

@@

[XXX To do: insert discussion of statistical mechanics and entropy here: This is why statistical mechanics works best on phase space (we have a uniform measure over q and p). Focusing on the density makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon and Gibbs entropies are coordinate invariant.]

Instead, of coordinate-independence, we take coordinate-invariance as a weaker necessary (but not sufficient) condition for fundamental physical significance. Being necessary for fundamental significance does not, however, make coordinate-invariance necessary for \textit{non-fundamental} physical significance. [XXX: disagree] Some quantities and properties are coordinate-variant yet still physically significant.\footnote{Indeed, many physically significant quantities are gauge-dependent, such as conjugate momentum and the Lagrangian (see Sections~\ref{gauge_momentum} and~\ref{gauge}). Insofar as gauge-dependency is analogous to being coordinate-variant, this supplies a further reason for denying that coordinate-invariance is a necessary condition for physical significance.} Our position here again places us at odds with others in the literature. For instance, North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}. However, some choices of unit-variables $q$ are not appropriately connected with physical measurement and hence are less physically meaningful [XXX to do: provide a clear example of this ---  not sure we can make this claim]. Similarly, we saw in Section~\ref{dependence} that some coordinate choices in Lagrangian mechanics are physically significant because they enable the property of point-wise comparison. This is related to the fact that, in Lagrangian mechanics, Cartesian coordinates are a physically significant coordinate choice because they decompose position and velocity variables as independent degrees of freedom. This independence does not hold generically across coordinate systems.\footnote{In contrast, canonical position and conjugate momentum variables always characterize separable/independent degrees of freedom in the Hamiltonian framework.} Finally, as the derivation in Section~\ref{deterministic} shows, a necessary condition for defining the symplectic structure is that we can express the density in a particular choice of units. If we did not have the physical ability to define the density in some units we would not be able to define this structure. This provides another instance of a physically significant but coordinate-dependent representational choice. Indeed, the next section shows that the conjugate momentum is

@@

The gauge-dependency of the conjugate momentum supplies further grounds for privileging the invariant phase space density. Insofar as coordinate-invariance is a plausible necessary condition for fundamental physical status, so is gauge-invariance. [XXX: check consistency with previous claim] By being gauge-variant, the conjugate momentum cannot be physically fundamental, although it remains physically significant (in the same way that coordinate-variant quantities can remain physically significant but not fundamental). [XXX: still, conjugate momentum is less physically significant because is not well defined experimentally]

Additionally, the gauge-dependency of conjugate momentum shows that not all state variables are on equal physical footing. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant. To see this, recall that the phase space density remains invariant upon changing the unit-variables $q$, inducing a state variable transformation on the conjugate momenta $p$ (which compensate by transforming covariantly). It is in this sense that the unit-variables $q$---rather than the conjugate momenta $p$---define the units of physical measurement, allowing us to assign numbers to physical quantities.\footnote{As discussed in Section~\ref{fundamental}, these units are physically significant despite not being coordinate-invariant.} Consequently, although symplectic mechanics allows arbitrary mixings of $q$'s and $p$'s that preserve the symplectic manifold, not all of these combinations are of equal physical significance. In general, the canonical position coordinates $q$ are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative $p$'s are not.

@@

Hamilton's principle of stationary action serves as a foundational principle for Lagrangian mechanics. According to this principle, infinitesimal variations of the action $ \mathscr{A}$ along the system's trajectory $\gamma $ equal zero: $\delta  \mathscr{A}[\gamma] = 0 $. Although central to Lagrangian mechanics, we can actually derive the stationary action principle using Hamiltonian mechanics. Our derivation simply assumes that 1) the system trajectory follows a deterministic and reversible evolution and 2) the system satisfies the motion equivalence condition. We present the derivation for the one-dimensional case---where it is most easily visualized---but it can be generalized using differential geometry.\footnote{For a similar argument illustrating this generalization, see \textcites[236--37, 243--45]{Arnold}.}
%our derivation seems very similar to Arnold chapter 9, section 45:C: ``the principle of least action in phase space''. Note that we briefly discussed Arnold on 11-23-20 meeting, around 12:30. 
% Note on the generalization: cannot use curl. Need to use a form that measures the number of states on a surface. The displacement field kills this form (eats the degenerate direction that does not form degrees of freedom).
% Something that I thought was a second assumption but turns out that this follows just from double role played by time variable: 2) the system's motion always advances in time (so that $dt/dt = 1$).

The setting is the extended phase space $(q, p, t)$, which is the standard phase space extended by the time variable. Knowing that the system is deterministic and reversible, it follows that the density of states is invariant. This means that possible trajectories of the system in the extended phase space are never pushed toward each other or pulled apart over time. In other words, the flux of trajectories through any closed surface in the extended phase space is always zero. Formally, the vector field $\vec{S} := (\frac{d q }{d t }, \frac{d p }{d t }, \frac{d t }{d t })$ represents the displacement of states along the trajectories. This field is divergenceless, $\nabla \cdot \vector{S} = 0$, because, given a region $U$ as many states will enter the region as exit. In particular, note that the displacement is tangent to the trajectory $\gamma [q, p, t] (t)$, in extended phase space, where the variable $t$ functions as both a coordinate and the parameter of the trajectory.

@@

(XXX to do: Note how the geometry is better understood in extended phase space, requiring the Hamiltonian picture. Note that this applies even when $\theta \cdot S$ cannot be expressed as a function of $q$ and $\dot q$. )

@@

Nonetheless, reformulating Hamiltonian mechanics within the extended phase space $(q_i, p_i, t, E)$ provides one way to rectify these deficiencies. At root, the time-dependence of the Hamiltonian stems from the dual role of time as both a coordinate and a time-like parameter within the usual Hamiltonian framework. Yet, differential geometry demands that we separate these two roles. We thereby add a temporal degree of freedom $(t, E) $, expanding the phase space degrees of freedom by one (recall that we previously had one degree of freedom for every canonical coordinate pair $q_i, p_i$). Next, we define the phase space evolution over an evolution parameter $s $. This yields two more equations of motion: $dt/ds = -\partial \mathcal{H}/ \partial E$ and $dE/ds = -\partial \mathcal{H}/ \partial t$, where $\mathcal{H}(q_i, p_i, t, E)$ is the extended Hamiltonian. The extended Hamiltonian not only acts as a generator of the evolution parameter $s$, but it also provides us with the \textit{Hamiltonian constraint} $\mathcal{H}=0$ [XXX to do: provide a reference for extened phase space formulation. And check this discussion of the `Hamiltonian constraint']. The Hamiltonian constraint provides the relationship between the energy and the other state variables. Crucially, the extended Hamiltonian---unlike the ordinary Hamiltonian---is coordinate-invariant, even with respect to time-dependent coordinate transformations. For example, if we define the affine parameter as the proper time, then the extended Hamiltonian is proportional to the mass (which is manifestly coordinate-invariant). This provides a clear physical meaning for the extended Hamiltonian. We have now considered many conceptual and physical advantages of the Hamiltonian formalism. In the next section, we rebut Curiel's \parencites*[]{Curiel} arguments against privileging Hamiltonian mechanics. 
%%Arnold uses a related extended phase space $(q_i, p_i, t)$ in his chapter 9, but doesn't provide a general discussion of it. 
%Discussion by GC of extended phase space: https://tinyurl.com/4tlbej76
% Is the Hamiltonian constraint for the extended phase space the same as the discussion in this Wikipedia page: https://en.wikipedia.org/wiki/Hamiltonian_constraint

@@

Whereas we can easily interpret non-hyper-regular Hamiltonians, the same cannot be said for all non-hyper-regular Lagrangians. For example, consider the case where the Lagrangian $L$ equals zero. This Lagrangian gives rise to a Hamiltonian that also equals zero. By Hamilton's equations, the derivatives of the position and momenta are zero, indicating that this system is static. [XXX: would remove mention of Hamiltonian] For this ``system," all paths have the same action, so all paths are possible. This gives an initial probability distribution over the states, but it does not specify how this initial distribution will evolve into a future distribution of states. In other words, the Lagrangian alone does not supply a probability distribution on the possible evolutions. [XXX: I would say, it does not tell us how an initial probability would evolve in time] In contrast, the Hamiltonian still defines a unique solution (i.e. a unique evolution in phase space) even when outside the hyper-regular domain (i.e. when the Hessian is zero).\footnote{In a footnote, Barrett conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" \parencites*[1179]{Barrett2}.} Indeed, Curiel acknowledges that only hyper-regular Lagrangians have unique solutions whereas Hamiltonian systems have unique solutions even outside of the hyper-regular domain \parencites*[291, 308]{Curiel}. Curiel argues that this is an advantage of Lagrangian mechanics, but it generates challenges for understanding non-hyper-regular Lagrangians.\footnote{This challenge is particularly pressing for systems that are everywhere non-regular. In the case where a system is non-regular only at certain points, we can decompose the system into hyper-regular subsystems, each with its own Lagrangian (or Hamiltonian).}
%The non-hyper-regular points form a boundary that we disregard.

>>>>

In the continuous case, $\mathcal{S}$ is a manifold, and we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. We will call a set of quantities $\xi^a$ \textit{state variables} provided that they fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates' and $\xi$ is a coordinate map. We will have multiple manifolds, so to avoid confusion, we will call the coordinates of the state space `state variables'. As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} In other words, $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument. 
% A state variable change $\hat{\xi}$ represents the same state $s$ by a different set of state variables $\hat{\xi}^b$.

Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$ (i.e. the choice of state variable map $\xi$). Hence, the density $\rho_{c, \xi}$ transforms as a scalar under state variable changes.\footnote{For two state variable maps $\xi$ and $\hat{\xi}$, we have $\rho_c(s) = (\rho_c \circ \xi^{-1} \circ \xi)(s) = (\rho_c \circ \hat{\xi}^{-1} \circ \hat{\xi})(s)$, which entails that $\rho_c(s) = \rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Here, $\xi(s) \equiv \xi^a$ and $\hat{\xi}(s) \equiv \hat{\xi}^b$.} Likewise, the integral $\int_U \rho_c(s) d\mathcal{S} $ is independent of the choice of state variables, so $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$. By the change of variables formula, we also know that $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a= \int_{\hat{\xi}(U)} \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$. Here, $ \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ is the determinant of the Jacobian for this change of state variables. Combining these two equations, we see that $\int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b = \int_{\hat{\xi}(U)} \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_{c, \hat{\xi}} (\hat{\xi}^b)d \hat{\xi}^b$, entailing that the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless. We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the classical phase space must be at least a symplectic manifold. 
%footnote{Indeed, using the Lebesgue measure $\lambda$ on $\mathbb{R}^n$---and assuming that $\xi^{-1}$: $\mathbb{R}^n \to \mathcal{S}$ is measurable---we can define the measure $d\mathcal{S}$ by the pushforward measure $\xi^{-1}_{\ast}(\lambda)(U) \equiv \lambda(\xi(U))$.} 
%In other words, while we can change the units used by the state variables $\xi^a$ to identify the particle states, the units for $\rho^{\ast}_c$ must remain the same so that the density does not change its numeric value

@@

The units we choose for $\omega$ are really the units we choose to count states over an independent degree of freedom. This agrees with the practice in statistical mechanics. The density $\rho_{c, \xi}$, when normalized, is given in terms of the inverse of the units for $\omega$. In accordance with the standard conventions in quantum mechanics and statistical mechanics, the $k_i$ are given in terms of inverse units of the respective $q^i$, such that $dq^i \wedge dk_i$ is a pure number. We then use $\hbar$ to define the units for state counting and therefore set $\omega = \hbar dq^i \wedge dk_i = dq^i \wedge dp_i$ where $p_i = \hbar k_i$.
% % the `motivation' section of the wiki page on symplectic manifolds provides a different style of derivation for the 2-form being non-degenerate and closed.

@@

This completes our central argument for privileging Hamiltonian mechanics over Lagrangian mechanics. We have seen that Lagrangian systems are a proper subset of Hamiltonian systems, specifically those Hamiltonian systems that also satisfy the motion equivalence condition. In the remainder of this essay, we provide further reasons for privileging Hamiltonian mechanics. The next section describes how Hamiltonian mechanics provides the natural setting for the invariant density, which characterizes how a classical object's parts are distributed. Section~\ref{fundamental} argues that the invariant phase space density is more fundamental than the state variables, since conjugate momentum is a gauge-dependent quantity. We then turn in Section~\ref{Lagrangian} to provide a geometric interpretation of the principle of stationary action within Hamiltonian mechanics. Our construction makes manifest the gauge redundancy of the Lagrangian, providing a further reason to privilege the Hamiltonian. Finally, we respond to Curiel's arguments against the fundamentality of Hamiltonian mechanics.

@@

[XXX To do: insert discussion of statistical mechanics and entropy here: statistical mechanics works best on phase space because phase space gives us a uniform measure over q and p. Focusing on the density makes the fundamental role that phase space plays in statistical mechanics more salient/perspicuous. Areas/volumes are measures of the number of states. They are coordinate invariant. Also the Shannon and Gibbs entropies are coordinate invariant.]\footnote{For further discussion of the importance of phase space for statistical mechanics and entropy, see \textcites{shpmp} [omit this for review?].}

Instead, of coordinate-independence, we take coordinate-invariance as a weaker necessary (but not sufficient) condition for fundamental physical significance. Being necessary for fundamental significance does not, however, make coordinate-invariance necessary for \textit{non-fundamental} physical significance. Many representational choices are coordinate-variant yet still physically significant.\footnote{Indeed, many physically significant quantities are gauge-dependent, such as conjugate momentum and the Lagrangian (see Sections~\ref{gauge_momentum} and~\ref{gauge}). Insofar as gauge-dependency is analogous to being coordinate-variant, this supplies a further reason for denying that coordinate-invariance is a necessary condition for physical significance.} Our position here again places us at odds with others in the literature. For instance, North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}. However, some choices of coordinates are particularly well-suited to certain problems, such as the use of action--angle variables for integrable systems, or spherical coordinates for systems with spherical symmetry. Similarly, we saw in Section~\ref{dependence} that some coordinate choices in Lagrangian mechanics are physically significant because they enable the property of point-wise comparison. This is related to the fact that, in Lagrangian mechanics, Cartesian coordinates are a physically significant coordinate choice because they decompose position and velocity variables as independent degrees of freedom. This independence does not hold generically across coordinate systems.\footnote{In contrast, canonical position and conjugate momentum variables always characterize separable/independent degrees of freedom in the Hamiltonian framework.} Finally, as the derivation in Section~\ref{deterministic} shows, a necessary condition for defining the symplectic structure is that we can express the density in a particular choice of units. If we did not have the physical ability to define the density in some units we would not be able to define this structure. This provides another instance of a physically significant but coordinate-dependent representational choice. Similarly, the next section shows that the conjugate momentum is gauge-dependent, but presumably still physically significant.

@@

The gauge-dependency of the conjugate momentum supplies further grounds for privileging the invariant phase space density. Insofar as coordinate-invariance is a plausible necessary condition for fundamental physical status, so is gauge-invariance. By being gauge-dependent, the conjugate momentum cannot be physically fundamental, although it remains physically significant (in the same way that coordinate-variant quantities can remain physically significant but not fundamental).
%[XXX: still, conjugate momentum is less physically significant because is not well defined experimentally]

Additionally, the gauge-dependency of conjugate momentum shows that not all state variables are on equal physical footing. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant. To see this, recall that the phase space density remains invariant upon changing the unit-variables $q$, inducing a state variable transformation on the conjugate momenta $p$ (which compensate by transforming covariantly). It is in this sense that the unit-variables $q$---rather than the conjugate momenta $p$---define the units of physical measurement, allowing us to assign numbers to physical quantities.\footnote{As discussed in Section~\ref{fundamental}, these units are physically significant despite not being coordinate-invariant.} Consequently, although symplectic mechanics allows arbitrary mixings of $q$'s and $p$'s that preserve the symplectic manifold, not all of these combinations are of equal physical significance. In general, the canonical position coordinates $q$ are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative $p$'s are not. Unlike kinetic momentum, conjugate momentum is not an experimentally measurable quantity, because its specification involves a gauge choice.

@@

Hamilton's principle of stationary action serves as a foundational principle for Lagrangian mechanics. According to this principle, infinitesimal variations of the action $ \mathscr{A}$ along the system's trajectory $\gamma $ equal zero: $\delta  \mathscr{A}[\gamma] = 0 $. Although central to Lagrangian mechanics, we can actually derive the stationary action principle using Hamiltonian mechanics. Our derivation simply assumes that 1) the system trajectory follows a deterministic and reversible evolution and 2) the system satisfies the motion equivalence condition. We present the derivation for the one-dimensional case---where it is most easily visualized---but it can be generalized using differential geometry.\footnote{For a similar argument, see \textcites[236--37, 243--45]{Arnold}.}
%our derivation seems very similar to Arnold chapter 9, section 45:C: ``the principle of least action in phase space''. Note that we briefly discussed Arnold on 11-23-20 meeting, around 12:30. 
% Note on the generalization: cannot use curl. Need to use a form that measures the number of states on a surface. The displacement field kills this form (eats the degenerate direction that does not form degrees of freedom).
% Something that I thought was a second assumption but turns out that this follows just from double role played by time variable: 2) the system's motion always advances in time (so that $dt/dt = 1$).

The derivation uses the extended phase space $(q, p, t)$, i.e. the standard phase space extended by the time variable. Knowing that the system is deterministic and reversible, it follows that the density of states is invariant. This means that possible trajectories of the system in the extended phase space are never pushed toward each other or pulled apart over time. In other words, the flux of trajectories through any closed surface in the extended phase space is always zero. Formally, the vector field $\vec{S} := (\frac{d q }{d t }, \frac{d p }{d t }, \frac{d t }{d t })$ represents the displacement of states along the trajectories. This field is divergenceless, $\nabla \cdot \vector{S} = 0$, because given a region $U$ as many states will enter the region as exit. In particular, note that the displacement is tangent to the trajectory $\gamma [q, p, t] (t)$ in extended phase space, where the variable $t$ functions as both a coordinate and the evolution parameter.

@@

%(XXX to do (might no longer be necessary given earlier and later remarks) Note how the geometry is better understood in extended phase space, requiring the Hamiltonian picture. Note that this applies even when $\theta \cdot S$ cannot be expressed as a function of $q$ and $\dot q$. )

@@

Nonetheless, reformulating Hamiltonian mechanics within the extended phase space $(q_i, p_i, t, E)$ provides one way to rectify these deficiencies.\footnote{For a detailed discussion of the extended phase space formalism, see \textcites[]{Struckmeier}.} At root, the time-dependence of the Hamiltonian stems from the dual role of time as both a coordinate and a time-like parameter within the usual Hamiltonian framework. Yet, differential geometry demands that we separate these two roles. We thereby add a temporal degree of freedom $(t, E) $, expanding the phase space degrees of freedom by one (recall that we previously had one degree of freedom for every canonical coordinate pair $q_i, p_i$). Next, we define the phase space evolution over an evolution parameter $s $. This yields two more equations of motion: $dt/ds = -\partial \mathcal{H}/ \partial E$ and $dE/ds = -\partial \mathcal{H}/ \partial t$, where $\mathcal{H}(q_i, p_i, t, E)$ is the extended Hamiltonian. The extended Hamiltonian not only acts as a generator of the evolution parameter $s$, but it also provides us with the \textit{Hamiltonian constraint} $\mathcal{H}=0$. The Hamiltonian constraint provides the relationship between the energy and the other state variables. Crucially, the extended Hamiltonian---unlike the ordinary Hamiltonian---is coordinate-invariant, even with respect to time-dependent coordinate transformations. For example, if we define the evolution parameter as the proper time, then the extended Hamiltonian is proportional to the mass (which is manifestly coordinate-invariant). This provides a clear physical meaning for the extended Hamiltonian. We have now considered many conceptual and physical advantages of the Hamiltonian formalism. In the next section, we rebut Curiel's \parencites*[]{Curiel} arguments against privileging Hamiltonian mechanics. 
%%Arnold uses a related extended phase space $(q_i, p_i, t)$ in his chapter 9, but doesn't provide a general discussion of it. 
%Discussion by GC of extended phase space: https://tinyurl.com/4tlbej76
% Is the Hamiltonian constraint for the extended phase space the same as the discussion in this Wikipedia page: https://en.wikipedia.org/wiki/Hamiltonian_constraint

@@

Whereas we can easily interpret non-hyper-regular Hamiltonians, the same cannot be said for all non-hyper-regular Lagrangians. For example, consider the case where the Lagrangian $L$ equals zero. For this ``system," all paths have the same action, so all paths are possible. Given an initial probability distribution over states, the Lagrangian fails to specify how this distribution will evolve in time. In contrast, a Hamiltonian still defines a unique solution (i.e. a unique evolution in phase space) even when outside the hyper-regular domain (i.e. when the Hessian is zero).\footnote{In a footnote, Barrett conjectures that outside the hyper-regular domain, Hamiltonian and Lagrangian formulations ``will be inequivalent according to any reasonable standard of equivalence. There is no natural way to translate a non-hyper-regular model of the one theory into a model of the other" \parencites*[1179]{Barrett2}.} For instance, a zero Hamiltonian describes a static system (by Hamilton's equations, the derivatives of the position and momenta are zero). Indeed, Curiel acknowledges that only hyper-regular Lagrangians have unique solutions whereas Hamiltonian systems have unique solutions even outside of the hyper-regular domain \parencites*[291, 308]{Curiel}. Curiel argues that this is an advantage of Lagrangian mechanics, but it generates challenges for understanding non-hyper-regular Lagrangians.\footnote{This challenge is particularly pressing for systems that are everywhere non-regular. In the case where a system is non-regular only at certain points, we can decompose the system into hyper-regular subsystems, each with its own Lagrangian (or Hamiltonian).}
%The non-hyper-regular points form a boundary that we disregard.
